Supplementary material: The increasing cost of happiness

### 2021-06-10 src/draft-SSM1.Rmd
Updates to text and Table S4. Closes issue #14 and issue #20

Subjective wellbeing is not a unitary entity [@diener2017findings]; studies typically distinguish between life satisfaction, the cognitive appraisal of one’s own accomplishments, and affective wellbeing, one’s prevailing affective state, emotional mood, or everyday experience of happiness [e.g., @kettlewell2020differential]. Kahneman & Deaton [-@kahneman2010high] showed that affective wellbeing increased with household income up to a point (USD75,000), but after that, increasing income had little further effect. Conversely cognitive wellbeing continued to increase with income beyond USD75,000. Indeed, the difference between the two questions: “How satisfied are you with your life?” and “How happy are you these days?” has been identified as a crucial mediating factor in a meta-analysis of 111 studies on income and wellbeing [@howell2008relation; also @veenhoven2006rising]. Studies since then have also shown that while both affective and cognitive wellbeing increase with income, the affective returns are a decreasing function of income [@killingsworth2021experienced; @sacks2012new].

While studies agree that cognitive wellbeing tends to increase linearly, or log-linearly, with income [e.g., @stevenson2013subjective; but see @jebb2018happiness], the exact form of the functional relationship between affective wellbeing and income is contested: some studies show the presence of a changepoint after which increasing income has little further effect [@kahneman2010high; @jebb2018happiness]; other studies have found that affective wellbeing is a constant (albiet decreasing) log-linear function of income [@killingsworth2021experienced; @sacks2012newl]; and indeed at least one recent study has indicated a constant linear relationship [@twenge2020expanding]. Few studies have formally tested the functional form of the relationship, relying instead upon visual comparison after binning the data over income deciles or quintiles. In this regard, arbitrary decisions about bin width may have salient effects on the functional form, which in turn will influence the inferences drawn. For this reason, a formal comparison between linear, piecewise-linear and log-linear functions in the relationship between affective wellbeing and income (unbinned) is warranted. More fundementally, the existence of a change point in the relationship between income and affective wellbeing would reveal an unacknowledged source of inequality in the distribution of wellbeing (i.e., happiness) in the economy. For instance, a change point of USD75,000 in 2008 was substantially more than the US median income of USD52,000 in the same year, indicating that the affective wellbeing of the poorest majority of the US population was tied to marginal changes in income while the affective wellbeing of a richer minority was not. Thus a change point would represent the dollar value up to which income drives inequities in the distribution of affective wellbeing, such that a lower valued change point represents a more equitable distribution of affective wellbeing in the economy.

@@

The evidence from model selection revealed the piecewise-linear fit of _affective_-, but not _cognitive_ wellbeing, was superior to a log-linear or linear fit in each year (Supplementary Materials, Figure S1). Overall the model comparison suggested that cognitive- and affective wellbeing have distinct relationships with household income; cognitive wellbeing tends to increase linearly with income, while a change point exists in the relationship between affective wellbeing and household income.

@@

We confirmed previous findings [e.g., @howell2008relation; @kahneman2010high], that the relationship between both types of subjective wellbeing with household income was positive but quite different: Cognitive wellbeing increased linearly with income, while affective wellbeing increased rapidly up to a point after which higher levels of income were associated with less improvement. Our novel finding was that this change point in the relationship between household income and affective wellbeing increased faster than both inflation and median household income between 2001 and 2019. For the first time we have shown there has been a temporal shift in the change point between income and affective wellbeing over the 19 year period, such that affective wellbeing has become more dependent on income for more people - especially the poor and middle-class.

>>>>

Subjective wellbeing is not a unitary entity [@diener2017findings]; studies typically distinguish between life satisfaction, the cognitive appraisal of one’s own accomplishments, and affective wellbeing, one’s prevailing affective state, emotional mood, or everyday experience of happiness [e.g., @kettlewell2020differential]. While measures of these distinct concepts are correlated, the distinction seems to be critical in understanding the impact of income on wellbeing [@howell2008relation]. For instance, we have recently observed that a major financial gain substantially improved cognitive wellbeing for individuals without much impact on their affective wellbeing [@kettlewell2020differential]. In a now classic paper, Kahneman & Deaton [-@kahneman2010high] showed that affective wellbeing increased with log household income up to a point (USD75,000), but after that, increasing income had no further effect (i.e., "satiation"). Conversely cognitive wellbeing continued to increase with (log) income beyond USD75,000. Studies since then have confirmed that while both affective and cognitive wellbeing increase with income, income is more strongly related to cognitive- than affective-wellbeing [@diener2013rising; @diener2017findings; @howell2008relation].

Fundementally, a difference in slope between affective wellbeing and income for the rich and poor represents an unacknowledged source of inequality in the distribution of wellbeing (i.e., happiness) in the economy. For instance, the "satiety" point of USD75,000 in 2008 was substantially more than the US median income of USD52,000 in the same year, indicating that the affective wellbeing of the poorest majority of the US population was tied to marginal changes in income while the affective wellbeing of a richer minority was not. While much commentary has focused on the zero slope at high income levels reported by Kahneman & Diener ("satiety" point), any decrease in slope for the richer minority would also represent a relatively unequal distribution of happiness among the poor compared to the rich. Thus, even a non-zero changepoint indicates that income is tied to inequities in the distribution of affective wellbeing in the economy. Conversely, a more equitable distribution of wellbeing in the economy would correspond to a lower valued change point.

While studies agree that affective wellbeing tends to increase with income, the exact form of the functional relationship between affective wellbeing and income is contested: some studies show the presence of a zero-slope changepoint after which increasing income has no further effect [@kahneman2010high; @jebb2018happiness]; other studies have found that affective wellbeing is a constant (albiet decreasing) log-linear function of income [@killingsworth2021experienced; @sacks2012new; @twenge2020expanding]. Few studies have formally tested the functional form of the relationship, relying instead upon binning the data (i.e., over income deciles or quintiles) and comparing adjacent bins either visually or with null hypothesis tests which do not provide evidence for a changepoint. In this regard, arbitrary choices about bin width can have salient effects on the function form, which in turn will influence the inferences drawn. And no studies to our knowledge have ever tested the presence of a changepoint between wellbeing and income using standard modelling approaches. For this reason, a formal comparison between log-linear, linear and a changepoint function (piecewise-linear) in the relationship between affective wellbeing and income is warranted.

@@

The evidence from model selection revealed the piecewise-linear fit of _affective_-, but not _cognitive_ wellbeing, was superior to a log-linear or linear fit in each year (Supplementary Materials, Figure S1). Overall the model comparison suggested that affective- and cognitive wellbeing have distinct relationships with household income; affective wellbeing increases with income more rapidly at lower household income levels than higher income levels with a distinct changepoint in each year, while cognitive wellbeing tends to increase with income, most likely in a linear fashion. Others have also noted cognitive wellbeing (life satisfaction) has a stronger linear relationship with income, than affective wellbeing and income [@howell2008relation], and the results of our formal comparison provides some support for this notion.

@@

We found the marginal effect of household income on affective- and cognitive-wellbeing was positive but quite different: Affective wellbeing increased rapidly up to a point after which higher levels of income were associated with less improvement, while cognitive wellbeing tended to increase linearly or constantly with income. Other studies have also reported that income has distinct effects on cognitive and affective wellbeing [@howell2008relation; @kahneman2010high], with a change point in affective wellbeing after which increasing income has little further effect. Our novel finding was revealing the change point increased faster than both inflation and median household income between 2001 and 2019. For the first time we have shown there has been a temporal shift in the change point between income and affective wellbeing over the 19 year period, such that affective wellbeing has become more dependent on income for more people - especially the poor and middle-class.

### 2021-06-10 src/supplementary-SSM1.Rmd
Updates to text and Table S4. Closes issue #14 and issue #20

The difference in WAIC scores for each model revealed the piecewise-linear fits of affective wellbeing, but not cognitive wellbeing, provided superior out-of-sample accuracy over a log-linear or linear fit in each year. Figure S1 shows the differences in WAIC scores clearly distinguished the advantage of the piecewise-linear fits for affective wellbeing with most 95% intervals entirely below zero for each year (exceptions occurred for the piecewise vs. log fit in 2004, 2006 and 2008, where the piecewise fit was still superior on average). By contrast, the difference in WAIC scores for the cognitive wellbeing models was not as clearly distinguished from zero for most years, since the 95% intervals of the WAIC differences with either the log-linear fit or the linear fit overlapped zero in most years. Indeed in the years between 2002 to 2010 the linear fit was superior to the piecewise-linear fit on average, with the mean WAIC difference above zero (positive). Overall the evidence suggested that affective wellbeing has a piecewise relationship with household income; implying the existence of a change point or difference in slopes at different levels of household income. Evidence of the effect of income on cognitive wellbeing, on the other hand, was more equivocal, albeit with some indication of a linear relationship in the early part of the 21st C. Others have also noted a linear relationship between cognitive wellbeing (life satisfaction) and income [@howell2008relation; @stevenson2013subjective], and the results of our formal comparison provides some support for this notion.

@@

##### Table S4. In-sample fit statistics (adj. R-squared)  
```{r table_s4}
model_summaries_all %>%
  select(model, Year = year, adj.r.squared) %>%
  spread(model, adj.r.squared) %>%
  flextable() %>%
  colformat_double(j = 2:7, digits = 3) %>%
  set_header_labels(
    lin = "Linear", 
    linc = "Linear\n(conditional)",
    log = "Log-linear",
    logc = "Log-linear\n(conditional)",
    pw = "Piecewise",
    pwc = "Piecewise\n(conditional)") %>%
  valign(i = 1, valign = "top", part = "header") %>%
  fontsize(size = 9.5, part = "all") %>%
  width(j = 2:7, width = 1) %>%
  width(j = 1, width = .5)
```

@@

A comparison of the relative differences in R^2^ values between models indicates the piecewise model produced a slightly higher in-sample fit over the other two models, while the linear model tended to have the lowest R^2^. The log-linear fit varied between both over the 19 years. As expected, adding covariates increased the R-squared in each case, but it did not change the relative difference between models.

>>>>

The difference in WAIC scores for each model revealed the piecewise-linear fits of affective wellbeing, but not cognitive wellbeing, provided superior out-of-sample accuracy over a log-linear or linear fit in each year. Figure S1 shows the differences in WAIC scores clearly distinguished the advantage of the piecewise-linear fits for affective wellbeing with most 95% intervals entirely below zero for each year (exceptions occurred for the piecewise vs. log fit in 2004, 2006 and 2008, where the piecewise fit was still superior on average). By contrast, the difference in WAIC scores for the cognitive wellbeing models was not as clearly distinguished from zero for most years, since the 95% intervals of the WAIC differences with either the log-linear fit or the linear fit overlapped zero in most years. Indeed in the years between 2002 to 2010 the linear fit was superior to the piecewise-linear fit on average, with the mean WAIC difference above zero (positive). Overall the evidence suggested that affective wellbeing has a piecewise relationship with household income; implying the existence of a change point or difference in slopes at different levels of household income. Evidence of the effect of income on cognitive wellbeing, on the other hand, was more equivocal, albeit with some indication of a linear relationship in the early part of the 21st C.

@@

##### Table S4. In-sample fit statistics (adj. R-squared) for affective wellbeing and income 
```{r table_s4}
model_summaries_all %>%
  select(model, Year = year, adj.r.squared) %>%
  spread(model, adj.r.squared) %>%
  flextable() %>%
  colformat_double(j = 2:7, digits = 3) %>%
  set_header_labels(
    lin = "Linear", 
    linc = "Linear\n(conditional)",
    log = "Log-linear",
    logc = "Log-linear\n(conditional)",
    pw = "Piecewise",
    pwc = "Piecewise\n(conditional)") %>%
  valign(i = 1, valign = "top", part = "header") %>%
  fontsize(size = 9.5, part = "all") %>%
  width(j = 2:7, width = 1) %>%
  width(j = 1, width = .5)
```

@@

A comparison of the relative differences in adjusted R^2^ values between models indicates the piecewise model produced a slightly higher in-sample fit over the other two models, while the linear model tended to have the lowest adjusted R^2^. The log-linear fit varied between both over the 19 years. As expected, adding covariates increased the adjusted R-squared in each case, but it did not change the relative difference between models.
